---
title: "Gerrymandering in Alabama: Analysis Plan"
author: "Samuel Barnard"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: sentence
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../docs") })
nocite: '@*'
bibliography: "../../software.bib"
---

# Abstract

This is a study of gerrymandering in Alabama.
We well test three methods of shape-based compactness scores, assess the representativeness of districts based on prior presidential elections and race.
We will then extend prior studies by calculating representativeness of the convex hull of district polygons.

# Study Metadata

-   `Key words`: gerrymandering, compactness, convex hull, alabama, political representation Comma-separated list of keywords (tags) for searchability. Geographers often use one or two keywords each for: theory, geographic context, and methods.
-   `Subject`: Social and Behavioral Sciences: Geography: Geographic Information Sciences
-   `Date created`: 2025-02-17
-   `Date modified`: 2025-02-17
-   `Spatial Coverage`: Alabama OSM:[161950](https://www.openstreetmap.org/relation/161950)
-   `Spatial Resolution`: Census Block Groups
-   `Spatial Reference System`: EPSG: 4269, NAD 1983 geographic coordinate system
-   `Temporal Coverage`: 2020-2024 population and voting data
-   `Temporal Resolution`: Decennial census

# Study design

This is an original study based on literature on gerrymandering metrics.

This study is **exploratory** in design, with the goal of evaluating usefulness of a new gerrymandering metric based on the convex hull of a congressional district and representative capability inside the convex hull compared to the congressional district.

# Materials and procedure

## Computational environment

I plan on using...\
`groundhog()` for reproducible computational environments (consistent versions of R and its packages)\
`here()` for reproducible path names\
`tidyverse()` includes `dplyr()` for database-style data frames\
`sf()` provides support for spatial vector data implementing the OSGeo simple features standards we are accustomed to\
`stars()` spatial-temporal raster data in R\
`tmap()` thematic maps, including static maps or interactive leaflet maps\

```{r environment-setup, include = FALSE}
# record all the packages you are using here
# this includes any calls to library(), require(),
# and double colons such as here::i_am()
packages <- c("tidyverse", 
              "here", 
              "sf", 
              "tmap", 
              "tidycensus", 
              "knitr", 
              "lwgeom",
              "markdown",
              "htmltools"
              )

# force all conflicts to become errors
# if you load dplyr and use filter(), R has to guess whether you mean dplyr::filter() or stats::filter()
# the conflicted package forces you to be explicit about this
# disable at your own peril
# https://conflicted.r-lib.org/
require(conflicted)

# load and install required packages
# https://groundhogr.com/
if (!require(groundhog)) {
  install.packages("groundhog")
  require(groundhog)
}

# this date will be used to determine the versions of R and your packages
# it is best practice to keep R and its packages up to date
groundhog.day <- "2025-02-01"

# this replaces any library() or require() calls
groundhog.library(packages, groundhog.day)
# you may need to install a correct version of R
# you may need to respond OK in the console to permit groundhog to install packages
# you may need to restart R and rerun this code to load installed packages
# In RStudio, restart r with Session -> Restart Session

# record the R processing environment
# alternatively, use devtools::session_info() for better results
writeLines(
  capture.output(sessionInfo()),
  here("procedure", "environment", paste0("r-environment-", Sys.Date(), ".txt"))
)

# save package citations
knitr::write_bib(c(packages, "base"), file = here("software.bib"))

# set up default knitr parameters
# https://yihui.org/knitr/options/
knitr::opts_chunk$set(
  echo = FALSE, # Run code, show outputs (don't show code)
  fig.retina = 4,
  fig.width = 8,
  fig.path = paste0(here("results", "figures"), "/")
)
```

------------------------------------------------------------------------

## Data and variables

We plan on using data sources...

Set up districts file from `districts.gpkg`:

```{r}
districts_file <- here("data", "raw", "public", "alabama_dataset", "districts.gpkg")
st_layers(districts_file)
```

### Layers from `districts.gpkg`

#### `precincts20`

-   `Title`: Voting Precincts 2020
-   `Abstract`: Alabama voting data for 2020 elections by precinct.
-   `Spatial Coverage`: Alabama OSM:[161950](https://www.openstreetmap.org/relation/161950)
-   `Spatial Resolution`: voting precincts
-   `Spatial Reference System`: EPSG: 4269, NAD 1983 geographic coordinate system
-   `Temporal Coverage`: voting precincts used for tabulating the 2020 election
-   `Temporal Resolution`: annual election (2020)
-   `Lineage`: Saved as geopackage format. Processing prior to download is explained in [validation report](al_vest_20_validation_report.pdf) and [readme](readme_al_vest_20.txt)
-   `Distribution`: Data available at [Redistricting Data Hub](https://redistrictingdatahub.org/dataset/vest-2020-alabama-precinct-and-election-results/) with free login.
-   `Constraints`: Permitted for noncommercial and nonpartisan use only. Copyright and use constraints explained [here](redistrictingdatahub_legal.txt)
-   `Data Quality`: State any planned quality assessment
-   `Variables`: For each variable, enter the following information. If you have two or more variables per data source, you may want to present this information in table form (shown below)
    -   `Label`: variable name as used in the data or code
    -   `Alias`: intuitive natural language name
    -   `Definition`: Short description or definition of the variable. Include measurement units in description.
    -   `Type`: data type, e.g. character string, integer, real
    -   `Accuracy`: e.g. uncertainty of measurements
    -   `Domain`: Expected range of Maximum and Minimum of numerical data, or codes or categories of nominal data, or reference to a standard codebook
    -   `Missing Data Value(s)`: Values used to represent missing data and frequency of missing data observations
    -   `Missing Data Frequency`: Frequency of missing data observations: not yet known for data to be collected

| Label | Alias | Definition | Type | Accuracy | Domain | Missing Data Value(s) | Missing Data Frequency |
|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
| VTDST20 | ... | Voting district ID | ... | ... | ... | ... | ... |
| GEOID20 | ... | Unique geographic ID | ... | ... | ... | ... | ... |
| G20PRETRU | ... | total votes for Trump in 2020 | ... | ... | ... | ... | ... |
| G20PREBID | ... | total votes for Biden in 2020 | ... | ... | ... | ... | ... |

**Load variables:**

```{r}
precincts20 <- st_read(districts_file, layer="precincts20")
```

#### `districts23`

-   `Title`: US Congressional Districts 2023
-   `Abstract`: Alabama congressional districts for the 2024 election.
-   `Spatial Coverage`: Alabama OSM:[161950](https://www.openstreetmap.org/relation/161950)
-   `Spatial Resolution`: congressional districts
-   `Spatial Reference System`: EPSG: 3857, NAD 1984 Web Mercator projection
-   `Temporal Coverage`: districts approved in 2023 for use in 2024.
-   `Temporal Resolution`:
-   `Lineage`: Loaded into QGIS as ArcGIS feaure service layer and saved in geopackage format. Extraneous data fields were removed and the `FIX GEOMETRIES` tool was used to correct geometry errors.
-   `Distribution`: Alabama State GIS [via ESRI feature service](https://services7.arcgis.com/jF2q3LPxL7PETdYk/arcgis/rest/services/2023_Court_Ordered_Congressional_Plan/FeatureServer/)
-   `Constraints`: Public Domain data free for use and redistribution.
-   `Data Quality`: State any planned quality assessment
-   `Variables`: For each variable, enter the following information. If you have two or more variables per data source, you may want to present this information in table form (shown below)
    -   `Label`: variable name as used in the data or code
    -   `Alias`: intuitive natural language name
    -   `Definition`: Short description or definition of the variable. Include measurement units in description.
    -   `Type`: data type, e.g. character string, integer, real
    -   `Accuracy`: e.g. uncertainty of measurements
    -   `Domain`: Expected range of Maximum and Minimum of numerical data, or codes or categories of nominal data, or reference to a standard codebook
    -   `Missing Data Value(s)`: Values used to represent missing data and frequency of missing data observations
    -   `Missing Data Frequency`: Frequency of missing data observations: not yet known for data to be collected

| Label | Alias | Definition | Type | Accuracy | Domain | Missing Data Value(s) | Missing Data Frequency |
|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
| DISTRICT | ... | US Congressional District Number | ... | ... | ... | ... | ... |
| POPULATION | ... | total population (2020 census) | ... | ... | ... | ... | ... |
| WHITE | ... | total white population (2020 census) | ... | ... | ... | ... | ... |
| BLACK | ... | total Black or African American population (2020 census) | ... | ... | ... | ... | ... |

**Load variables:**

```{r}
districts23 <- st_read(districts_file, layer="districts23")
```

**Map 2023 districts**

```{r}
tmap_mode(mode = "plot")

districts23map <- districts23 |> 
  tm_shape() +
  tm_polygons(fill_alpha = 0,
              col = "black") +
  tm_labels(text = "DISTRICT",
          col="black",
          bgcol = "white",
          bgcol_alpha = 0.5,
          on_surface = TRUE,
          just = c("center", "center")
          )

districts23map #render map
```

### `blockgroups2020`

-   `Title`: Block Groups 2020
-   `Abstract`: Vector polygon geopackage layer of Census tracts and demographic data.
-   `Spatial Coverage`: Alabama OSM:[161950](https://www.openstreetmap.org/relation/161950)
-   `Spatial Resolution`: census block groups
-   `Spatial Reference System`: EPSG: 4269, NAD 1983 geographic coordinate system
-   `Temporal Coverage`: 2020 census
-   `Temporal Resolution`: 10 year census (2020)
-   `Lineage`: Data downloaded from US Census API "pl" public law summary file using tidycensus in R
-   `Distribution`: US Census API
-   `Constraints`: Public Domain data free for use and redistribution.
-   `Data Quality`: State any planned quality assessment
-   `Variables`: For each variable, enter the following information. If you have two or more variables per data source, you may want to present this information in table form (shown below)
    -   `Label`: variable name as used in the data or code
    -   `Alias`: intuitive natural language name
    -   `Definition`: Short description or definition of the variable. Include measurement units in description.
    -   `Type`: data type, e.g. character string, integer, real
    -   `Accuracy`: e.g. uncertainty of measurements
    -   `Domain`: Expected range of Maximum and Minimum of numerical data, or codes or categories of nominal data, or reference to a standard codebook
    -   `Missing Data Value(s)`: Values used to represent missing data and frequency of missing data observations
    -   `Missing Data Frequency`: Frequency of missing data observations: not yet known for data to be collected

| Label | Alias | Definition | Type | Accuracy | Domain | Missing Data Value(s) | Missing Data Frequency |
|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
| GEOID | ... | code to uniquely identify tracts | ... | ... | ... | ... | ... |
| P4_001N | ... | total population, 18 years or older | ... | ... | ... | ... | ... |
| P4_006N | ... | total: not Hispanic or Latino, Population of one race, Black or African American alone, 18 years or older | ... | ... | ... | ... | ... |
| P5_003N | ... | Total institutionalized population in correctional facilities for adults, 18 years or older | ... | ... | ... | ... | ... |

**Load data:**

Acquire decennial census data in block groups using the `tidycensus` package.
First, query metadata for the `pl` public law data series.

```{r}
census_metadata_file <- here("data", "metadata", "census2020pl_vars.csv")
if(file.exists(census_metadata_file)){
  census2020pl_vars <- read.csv(census_metadata_file)
} else {
  census2020pl_vars <- load_variables(2020, "pl")
  write.csv(census2020pl_vars, here("data", "metadata", "census2020pl_vars.csv"))
}
```

The issue in the 2023 court cases on Alabama's gerrymandering was a racial gerrymander discriminating against people identifying as Black or African American.
Therefore, we will analyze people of voting age (18 or older) identifying as Black and or African as one race in any combination with other races.

This data is found in public law data series table `P3`.

Query table `P3` on `"race for the population 18 years and over"`.

```{r message=FALSE, warning=FALSE}
blockgroup_file <- here("data", "raw", "public", "block_groups.gpkg")

# if the data is already downloaded, just load it
# otherwise, query from the census and save
if(file.exists(blockgroup_file)){
  blockgroups <- st_read(blockgroup_file)
} else {
  blockgroups <- get_decennial(geography = "block group",
                               sumfile = "pl",
                               table = "P3",
                               year = 2020,
                               state = "Alabama",
                               output = "wide",
                               geometry = TRUE,
                               keep_geo_vars = TRUE)
  st_write(blockgroups, blockgroup_file)
}
```

## Prior observations

I have previously conducted an analysis with this data using QGIS to determine compactness along with race and party affiliation data.

However, I only conducted my analysis with an area-weighted re-aggregation approach, and did not incorporate convex hulls.

## Bias and threats to validity

"This study is explicitly an investigation to the modifiable areal unit problem. Aspects of the study are extremely sensitive to the combination of edge effects and scale, whereby complex borders formed by natural features, e.g. coastlines or rivers, vary greatly in perimeter depending on the scale of analysis. We hope that in part, this study establishes a method that is more robust (less sensitive) to the threats to validity caused by scale and edge effects in studies of gerrymandering and district shapes."

------------------------------------------------------------------------

## Data transformations

### Step 1:

`districts23` needs to be re-projected to EPSG:4269 NAD 1983 coordinate system using `st_transform()` for the purpose of geodesic analysis.

From here, we can calculate the percentage of population identifying as Black using `mutate()`.

```{r}
districts23 <- districts23 |> st_transform(4269) |> 
  mutate(pctBlack = round(BLACK / POPULATION * 100, 1))
```

### Step 2:

Census data (`blockgroups2020`) also needs to be re-projected from the WGS 1984 geographic coordinate system to the NAD 1983 geographic coordinate system.

```{r}
blockgroups <- st_transform(blockgroups, 4269)
```

### Step 3:

Find the total of people identifying as Black or African American as one race or any combination of multiple races.

#### a. First, make a list of all the variables inclusive of people identifying as Black or African American.

```{r}
baa_vars <- census2020pl_vars |> 
  dplyr::filter(str_detect(name, "P3"),
                str_detect(label, "Black")) |> 
  select(-concept)

baa_vars |> kable()
```

#### b. Next, calculate new columns.

`Black` is a sum of all 32 columns shown above, in which any of the racial categories by which someone identifies is Black or African American.\
`Total` is a copy of the population 18 years or over, variable `P3_001N`.\
`PctBlack` is calculated as `Black / Total * 100`\
`CheckPct` is calculated as the percentage of the population 18 years or older that is either white of one race only (`P3_003N`) or Black or African American as calculated above.
In Alabama, we can expect that this will be close to 100% for most block groups, and should never exceed 100%.

```{r}
blockgroups_calc <- blockgroups |> 
  rowwise() |> 
  mutate(Black = sum(c_across(all_of(baa_vars$name)))) |> 
  ungroup() |> 
  mutate(bgarea = st_area(geometry),
         Total = P3_001N,
         PctBlack = Black / Total * 100,
         CheckPct = (Black + P3_003N) / Total * 100
         ) |> 
  select(GEOID, bgarea, Black, Total, PctBlack, CheckPct)
```

#### c. Save the results in `blockgroups_calc.gpkg`

```{r}
st_write(blockgroups_calc, 
         here("data", "derived", "public", "blockgroups_calc.gpkg"),
         append=FALSE)
```

### Step 4:

Map the percentage of the population 18 or over that is Black or African American.

```{r}
tmap_mode(mode = "plot")
blkgrp_baa_map <- tm_shape(blockgroups_calc) + 
  tm_polygons(
    fill = "PctBlack",
    col_alpha = 0.2,
    lwd = 0.1,
    col = "grey90"
  )

blkgrp_baa_map
```

Map approved 2023 districts over the black population

```{r}
tmap_mode(mode = "view") #non-static map view

blkgrp_baa_map +
  districts23map
```

------------------------------------------------------------------------

## Analysis

### Approach 1: AWR

Use area weighted re-aggregation to estimate white and black voting age populations in block groups.

Why do this when `POPULATION`, `BLACK`, and `WHITE` variables are already in the table?
*First*, this is the total population, but we should care more about the voting age population.
*Second*, we may want to categorize and calculate `BLACK` differently from the state of Alabama.

It turns out that R optimizes the **first** dataset in a spatial query or overlay, with a spatial index, and not the second.
Therefore, add the more complex data to `st_intersection` first, and you’ll see remarkably different run times.

[Spatial indices in R (sf)](https://r-spatial.org/r/2017/06/22/spatial-index.html)

```{r}
districts23_estimates <- st_intersection(blockgroups_calc, districts23) |> 
  mutate(
    awTot = Total * as.numeric(st_area(geometry) / bgarea),
    awBlack = Black * as.numeric(st_area(geometry) / bgarea)
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT) |> 
  summarize(bgTotal = sum(awTot),
            bgBlack = sum(awBlack))

districts23_join_bg <- districts23 |> 
  left_join(districts23_estimates, by = "DISTRICT") |> 
  mutate(pctBlackbg = round(bgBlack / bgTotal * 100, 1))
```

Report results. Percentages of Black or African American people vary significantly from district to district.

```{r}
districts23_join_bg |> st_drop_geometry() |> kable()
```

### Approach 2: Convex hull

```{r}
districts23_estimates <- st_intersection(blockgroups_calc, st_convex_hull(districts23)) |> 
  mutate(
    awTot = Total * as.numeric(st_area(geometry) / bgarea),
    awBlack = Black * as.numeric(st_area(geometry) / bgarea)
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT) |> 
  summarize(chTotal = sum(awTot),
            chBlack = sum(awBlack))
```

Join convex hull estimates to districts with blockgroup estimates.

```{r}
districts23_join_ch <- districts23_join_bg |> 
  left_join(districts23_estimates, by = "DISTRICT") |> 
  mutate(pctBlackch = round(chBlack / chTotal * 100, 1),
         diffPct = pctBlackbg - pctBlackch,
         absdiffPct = abs(diffPct))
```

Calculate compactness scores based on:

1. the area and perimeter
2. the area and the area of the convex hull
3. the area and the area of the minimum bounding circle

Note: This code block takes some time to run due to the `st_minimum_bounding_circle` function.

Also, to knit, will we need to replace`st_perimeter()` with `st_length(st_cast(geom, "MULTILINESTRING"))`?

```{r}
districts23_results <- districts23_join_ch |> 
  mutate(
    darea = st_area(geom),
    dperim = st_length(st_cast(geom, "MULTILINESTRING")),
#    dperim2 = st_perimeter(geom),
    compact_shp = round( as.numeric((4 * pi * darea) / dperim^2), 2),
    compact_hull = round( as.numeric(darea / st_area(st_convex_hull(geom))), 2),
    compact_circ = round( as.numeric(darea / st_area(st_minimum_bounding_circle(geom))), 2)
  )
```

------------------------------------------------------------------------

# Results

## Visualization 1: Correlation matrix and small plots for gerrymandering indicators

```{r}
districts23_results_cor <- districts23_results |> 
  st_drop_geometry() |> 
  select(pctBlackbg,
         diffPct,
         absdiffPct,
         compact_shp,
         compact_hull,
         compact_circ)

districts23_results_cor |> cor() |> kable()
```

```{r}
districts23_results_cor |> pairs()
```

## Visualization 2: Plot with representational difference and compactness

This is a scatterplot with (absolute) difference in representation on x axis and compactness on y axis. This plot presents the three different compactness scores simultaneously with different colors.

```{r}
districts23_results_plot <- districts23_results |> 
  st_drop_geometry() |> 
  select(DISTRICT, pctBlack, absdiffPct, compact_shp, compact_hull, compact_circ) |> 
  pivot_longer(cols = starts_with("compact"))

districts23_results_plot |> ggplot() +
  aes(x = absdiffPct, y = value) +
  geom_smooth(method="lm", col = "grey30") +
  geom_label(aes(label = DISTRICT, fill = pctBlack)) +
  scale_fill_distiller(type = "div", palette = "PRGn") +

  facet_wrap(~name)
```
There is a positive relationship between minimum bounding circle compactness and absolute convex hull representational difference. There is a negative relationship between convex hull compactness and absolute convex hull representational difference. There is a negative relationship between shape compactness and absolute convex hull representational difference. 
Shape and convex hull exhibit a positive correlation.

Districts 1 and 2 are the least compact across all scores.

```{r}
tm_shape(districts23_results) +
  tm_polygons(fill = "pctBlackbg") +
  tm_text("DISTRICT")
```

# Discussion

Describe how the results are to be interpreted *vis a vis* each hypothesis or research question.

# Integrity Statement

Include an integrity statement - The authors of this preregistration state that they completed this preregistration to the best of their knowledge and that no other preregistration exists pertaining to the same hypotheses and research.
If a prior registration *does* exist, explain the rationale for revising the registration here.

# Acknowledgements

This analysis is based on work by Professor Joseph Holler of Middlebury College, and concepts and workflows from the course Open GIScience GEOG 0361. 

This report is based upon the template for Reproducible and Replicable Research in Human-Environment and Geographical Sciences, [DOI:[10.17605/OSF.IO/W29MQ](DOI:%5B10.17605/OSF.IO/W29MQ){.uri}](https://doi.org/10.17605/OSF.IO/W29MQ)

# References

Cheng, Joe, Carson Sievert, Barret Schloerke, Winston Chang, Yihui Xie, and Jeff Allen. 2024. Htmltools: Tools for HTML. https://github.com/rstudio/htmltools.
Müller, Kirill. 2020. Here: A Simpler Way to Find Your Files. https://here.r-lib.org/.
Pebesma, Edzer. 2018. “Simple Features for R: Standardized Support for Spatial Vector Data.” The R Journal 10 (1): 439–46. https://doi.org/10.32614/RJ-2018-009.
———. 2024a. Lwgeom: Bindings to Selected Liblwgeom Functions for Simple Features. https://r-spatial.github.io/lwgeom/.
———. 2024b. Sf: Simple Features for r. https://r-spatial.github.io/sf/.
Pebesma, Edzer, and Roger Bivand. 2023. Spatial Data Science: With applications in R. Chapman and Hall/CRC. https://doi.org/10.1201/9780429459016.
R Core Team. 2024. R: A Language and Environment for Statistical Computing. Vienna, Austria: R Foundation for Statistical Computing. https://www.R-project.org/.
Tennekes, Martijn. 2018. “tmap: Thematic Maps in R.” Journal of Statistical Software 84 (6): 1–39. https://doi.org/10.18637/jss.v084.i06.
———. 2025. Tmap: Thematic Maps. https://github.com/r-tmap/tmap.
Walker, Kyle, and Matt Herman. 2025. Tidycensus: Load US Census Boundary and Attribute Data as Tidyverse and Sf-Ready Data Frames. https://walker-data.com/tidycensus/.
Wickham, Hadley. 2023. Tidyverse: Easily Install and Load the Tidyverse. https://tidyverse.tidyverse.org.
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D’Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. “Welcome to the tidyverse.” Journal of Open Source Software 4 (43): 1686. https://doi.org/10.21105/joss.01686.
Xie, Yihui. 2014. “Knitr: A Comprehensive Tool for Reproducible Research in R.” In Implementing Reproducible Computational Research, edited by Victoria Stodden, Friedrich Leisch, and Roger D. Peng. Chapman; Hall/CRC.
———. 2015. Dynamic Documents with R and Knitr. 2nd ed. Boca Raton, Florida: Chapman; Hall/CRC. https://yihui.org/knitr/.
———. 2024. Knitr: A General-Purpose Package for Dynamic Report Generation in r. https://yihui.org/knitr/.
Xie, Yihui, JJ Allaire, and Jeffrey Horner. 2024. Markdown: Render Markdown with Commonmark. https://github.com/rstudio/markdown.
